Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-12 draft/draft.tex
minor changes
updated draft

% title, authors, affiliations
\title{Cohesiveness in microbial community coalescence}
\author[1,2,3]{Authors}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Other affiliations...}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
\date{}

@@

%\linenumbers

@@

We performed a new round of invasion experiments where each dominant invaded every
other community in isolation (\methodsref{competitions}). The dominant-invaded
communities were then propagated for seven serial transfers, and the final
community compositions were determined by 16S Illumina sequencing.
We compared the relative abundances reached by the dominants in these experiments
with the relative abundances of the same dominants in community coalescence
(i.e. when they invade with their cohorts).
\figref[b]{fig2} shows that, for communities assembled in citrate environments,
the cohorts had little effect and the dominants were able to invade with comparable
success regardless of being accompanied by them.
This was expected since we had seen that coalescence between citrate communities
tended to be determined by top-down co-selection.
However, glutamine communities displayed a different behavior.
A large number of dominants (roughly 50\%) that were not able to invade by themselves
(or could only do so at very low final relative abundances, below 0.1)
were able to reach much higher frequencies when they invaded together with their cohorts,
which indicates that positive bottom-up co-selection is common in our experiments.
In contrast, negative co-selection appears to be rare and no instances of it were
observed in our experiments.

@@

We ran a set of simulations of community coalescence to better understand the mechanisms
of ecological co-selection in our experiments. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}.
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we replicated our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were then allowed to stabilize through 20
growth-dilution cycles. Stable communities were then used to reproduce our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

@@

Understanding the mechanisms underlying the responses of microbial communities to invasions is
an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
Theory has suggested that communities may exhibit an emergent cohesiveness
\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
leading to members of the same community recruiting one another during
community-community invasions.
Our results provide direct experimental evidence of ecological co-selection in a large number
of community coalescence experiments,
and highlight the critical role played by the rarer, sub-dominant species in the generation
of community cohesiveness.
Our data suggests that the strength and direction of ecological co-selection is modulated
by the supplied primary carbon source,
which in turn shapes the cross-feeding network that dictates
the structure of our communities \cite{Goldford2018,Estrela2020}.
These observations,
together with previous experimental results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions arising from metabolic feedbacks between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Further work is required to clarify the relationship between metabolic interactions,
community structure, cohesiveness and co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.

@@

The authors wish to thank Joshua Goldford, Pankaj Mehta, Wenping Cui,
Robert Marsland and all members of the Sanchez laboratory
for many helpful discussions.
We also wish to express our gratitude to the Goodman laboratory at Yale
for technical help during the early stages of this project.
The funding for this work partly results from a Scialog Program
sponsored jointly by the Research Corporation for Science Advancement and
the Gordon and Betty Moore Foundation through grants to Yale University by the
Research Corporation and the Simons Foundation.

>>>>

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,2,*]{Juan Diaz-Colunga}
\author[1,2,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4,5]{Mikhail Tikhonov}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
Cambridge, MA, USA}
\affil[4]{Department of Applied Physics,
Stanford University, Stanford, CA, USA}
\affil[5]{Department of Physics,
Washington University, St. Louis, MO, USA}

\affil[$\dagger$]{\normalfont To whom correspondence should be addressed: alvaro.sanchez@yale.edu}
\affil[*]{\normalfont These authors contributed equally}
\date{}

@@

\linenumbers

@@

To test which of these situations was more common in our conditions,
we performed a new round of invasion experiments where each dominant invaded every
other community in isolation (\methodsref{competitions}). The dominant-invaded
communities were then propagated for seven serial transfers, and the final
community compositions were determined by 16S Illumina sequencing.
We compared the relative abundances reached by the dominants in these experiments
with the relative abundances of the same dominants in community coalescence
(i.e. when they invade with their cohorts).
\figref[b]{fig2} shows that, for communities assembled in citrate environments,
the cohorts had little effect and the dominants were able to invade with comparable
success regardless of being accompanied by them.
This was expected since we had seen that coalescence between citrate communities
tended to be determined by top-down co-selection.
However, glutamine communities displayed a different behavior.
A large number of dominants (roughly 50\%) that were not able to invade by themselves
(or could only do so at very low final relative abundances, below 0.1)
were able to reach much higher frequencies when they invaded together with their cohorts,
which indicates that positive bottom-up co-selection is common in our experiments.
In contrast, negative co-selection appears to be rare and no instances of it were
observed in our experiments.

@@

To better understand the mechanisms of ecological co-selection in our experiments,
we ran a set of simulations of community coalescence. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we replicated our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were then allowed to stabilize through 20
growth-dilution cycles. Stable communities were then used to reproduce our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

@@

Understanding the mechanisms underlying the responses of microbial communities to invasions is
an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
Theory has suggested that communities may exhibit an emergent cohesiveness
\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
leading to members of the same community recruiting one another during
community-community invasions.
Our results provide direct experimental evidence of ecological co-selection in a large number
of community coalescence experiments,
and highlight the critical role played by the rarer, sub-dominant species in the generation
of community cohesiveness.
Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic network that shapes the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
The architecture of this network is dependent on the supplied primary carbon source.
These observations,
together with previous experimental results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Further work will be necessary to clarify the relationship between
metabolic feedbacks, community cohesiveness, and
the strength and direction of ecological co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.

@@

The authors wish to thank Pankaj Mehta, Wenping Cui,
Robert Marsland and all members of the Sanchez laboratory
for many helpful discussions.
We also wish to express our gratitude to the Goodman laboratory at Yale
for technical help during the early stages of this project.
The funding for this work partly results from a Scialog Program
sponsored jointly by the Research Corporation for Science Advancement and
the Gordon and Betty Moore Foundation through grants to Yale University by the
Research Corporation and the Simons Foundation.
